cardinalR: Generating interesting high-dimensional data structures

### 2025-05-19 paper-cardinalR.tex
Fixed the issu of rendering the pdf

The package can be installed from CRAN using

\begin{verbatim}
install.packages("cardinalR")
\end{verbatim}

and the development version can be installed from GitHub

@@

\subsection{Web site}\label{web-site}

More documentation of the package can be found at the web site \url{https://jayanilakshika.github.io/cardinalR/}.

\subsection{Data sets}\label{data-sets}

The \texttt{cardinalR} package comes with several data sets that load with the package. These are described in Table \ref{tab:datasets-tb-pdf}.

\caption{\label{tab:datasets-tb-pdf}cardinalR data sets}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
data & explanation\\
\midrule
mobiusgau & Simulated data with a Mobius and a Gaussian in 4-D space.\\
mobiusgau\_tsne1 & The tSNE embedding with perplexity \$15\$ for mobiusgau.\\
mobiusgau\_tsne2 & The tSNE embedding with perplexity \$30\$ for mobiusgau.\\
mobiusgau\_tsne3 & The tSNE embedding with perplexity \$5\$ for mobiusgau.\\
mobiusgau\_umap1 & The UMAP embedding with number of neighbors \$15\$ and minimum distance \$0.1\$ for mobiusgau.\\
\addlinespace
mobiusgau\_umap2 & The UMAP embedding with number of neighbors \$30\$ and minimum distance \$0.08\$ for mobiusgau.\\
mobiusgau\_umap3 & The UMAP embedding with number of neighbors \$5\$ and minimum distance \$0.9\$ for mobiusgau.\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Functions}\label{functions}

@@

The \texttt{gen\_expbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) exponential branches in 2D, with optional noise dimensions to embed the structure in a \$p\$-dimensional space. These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

@@

\(X\_1 \sim \mathcal{U}(-2, 2)\)
\(X\_2 = \exp(\pm s\_i \cdot X\_1) + \epsilon\),

where:
\(\epsilon \sim \mathcal{U}(0, 0.1)\) adds local jitter,

\(s\_i \in [0.5, 2]\) is randomly sampled for each branch.

@@

\[
X\_j \sim \mathcal{N}(0, 0.1^2), \quad j = 3, \dots, p.
\]

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from \(X\_1 \sim \mathcal{U}(-2, 8)\$, \$X\_2 = 0.5 \cdot X\_1 + \epsilon\). The Branch 2 is generated from \(X\_1 \sim \mathcal{U}(-6, 2)\$, \$X\_2 = -0.5 \cdot X\_1 + \epsilon\). \(\epsilon \sim \mathcal{U}(0, 0.5)\).

Branches 3 to \(k\)** are added iteratively. Each starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches.

\(X\_1\) values span a short segment (\(x\_{start}\) to \(x\_{start} + 1\)),

\(X\_2 = s\_i \cdot (X\_1 - x\_{start}) + y\_{start} + \epsilon\) where \(s\_i\) is a sampled slope from a filtered set and \(\epsilon \sim \mathcal{U}(0, 0.2)\).

@@

\[
X\_j \sim \mathcal{N}(0, 0.05^2)\$ for \$j = 3, \dots, p.
\]

@@

Branch 1 is generated from \(X\_1 \sim \mathcal{U}(0, 1)\), and \(X\_2 = 0.1 \cdot X\_1 + 1 \cdot X\_1^2 + \epsilon\), where \(\epsilon \sim \mathcal{U}(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane.

Branch 2 is generated from \(X\_1 \sim \mathcal{U}(-1, 0)\), and \(X\_2 = 0.1 \cdot X\_1 - 2 \cdot X\_1^2 + \epsilon\), where \(\epsilon \sim \mathcal{U}(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane.

Branches \(3\) to \(k\)** are added iteratively. Each new branch begins at a randomly chosen point within a restricted horizontal band: \(X\_1 \in [-0.15, 0.15]\) (to ensure connectivity with earlier branches), spans a unit-length interval on \(X\_1\):\(X\_1 \in [x\_{\text{start}}, x\_{\text{start}} + 1]\), and has the structure:

@@

Here, \(s\_i\) is a sampled scale factor (slope-like term) controlling curvature, and \((x\_{\text{start}}, y\_{\text{start}})\) is the starting point from the existing structure.

@@

\[
X_j \sim \mathcal{N}(0, 0.05^2), \quad \text{for } j = 3, \dots, p.
\]

@@

To construct each branch, a unique or repeated pair of dimensions is selected from the \(\binom{p}{2}\) possible \(2\text{-}D\) combinations. If \(k \leq \binom{p}{2}\), combinations are sampled without replacement. If \(k > \binom{p}{2}\), additional pairs are sampled with replacement to reach \(k\) total branches. Each selected pair \((x\_{i1}, x\_{i2})\) defines the \(2\text{-}D\) plane for branch \(i\).

For branch \(i\), \(n\_i\) points are generated, where \(\sum\_{i=1}^k n\_i = n\). The structure follows:

\[
x_{i1} \sim \mathcal{U}(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim \mathcal{U}(0, 0.5),
\]

where \(s\_i\) is a scale factor controlling the slope. When \(k \leq \binom{p}{2}\), \(s\_i = 1\). When sampling with replacement, \(s\_i\) is drawn from the set \({1, 1.5, 2, \dots, 8}\) in increments of \(0.5\).

@@

\[
x_j \sim \mathcal{N}(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
\]

@@

Let \(\binom{p}{2}\) denote the number of unique \(2\text{-}D\) subspace combinations. When \(k \leq \binom{p}{2}\), \(k\) distinct subspace pairs \((x\_{i1}, x\_{i2})\) are sampled without replacement. Otherwise, combinations are selected with replacement to reach \(k\) total branches. For each branch \(i = 1, \dots, k\), a random scale factor \(s\_i\) is used to vary the curvature:

\begin{itemize}
\tightlist
\item
  If \(k \leq \binom{p}{2}\): \$s\_i = 1 \$
\item
  If \(k > \binom{p}{2}\): \(s\_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}\)
\end{itemize}

Each branch contains \(n\_i\) points such that \(\sum\_{i=1}^k n\_i = n\), where the vector \((n\_1, \dots, n\_k)\) is randomly drawn using the helper function \texttt{gen\_nsum()} to partition \(n\).

Within its assigned subspace \((x\_{i1}, x\_{i2})\), branch \(i\) is defined by:

\[
x_{i1} \sim \mathcal{U}(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1}^2 + \epsilon, \quad \epsilon \sim \mathcal{U}(0, 0.5)
\]

This forms a smooth downward-opening parabola in the plane defined by \((x\_{i1}, x\_{i2})\), with the degree of curvature controlled by \(s\_i\).

@@

Each variable \(X\_j\) represents integer grid coordinates (e.g., 1, 2, 3, \ldots) in dimension \(j\):

\[
X\_j \in {1, \dots, k\_j},\text{ where }\prod\_{j=1}^p k\_j \approx n.
\]

@@

Points are sampled independently from \(X\_1, X\_2, X\_3 \sim \mathcal{U}(-0.5, 0.5)\) of a cube of side length 1. This results in a uniform cloud of points filling the unit cube in \(3\text{-}D\) space.

@@

\[
X\_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
\]

@@

The initial points are drawn from a uniform cube centered at the origin: \(X\_1, X\_2, X\_3 \sim \mathcal{U}(-0.5, 0.5)\).

@@

Remove if \(\sqrt{X\_1^2 + X\_2^2 + X\_3^2} \leq 0.5\).

@@

\[
X\_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
\].

@@

\[
X_i \sim \mathcal{N}_p(\boldsymbol{0}, \Sigma)
\]

@@

\(\text{scale}_i \sim \mathcal{U}(-10, 10)\): randomly chosen direction and stretch.

\(\text{shift}_i \sim \mathcal{U}(-300, 300)\): large offset to spread features apart.

\(\epsilon \sim \mathcal{N}(0, 0.03 \cdot n)\): Gaussian noise scaled to dataset size to introduce mild irregularity.

@@

\[
X\_1 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \cos(\theta),
\]
\[
X\_2 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \sin(\theta),
\]
\[
X\_3 = \frac{w}{2} \sin(\theta/2).
\]

@@

\[
x_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
\]

@@

The curve is constructed by drawing uniformly spaced inputs and applying a second-degree polynomial transformation. Let \(X\_1 \sim \mathcal{U}(\text{range}[1], \text{range}[2])\) be the independent variable. A raw polynomial basis of degree 2 is used to form \(X\_2 = X\_1 - X\_1^2 + \varepsilon\_2\), where \(\varepsilon\_2 \sim \mathcal{U}(0, 0.5)\).

@@

The shape is generated using a third-degree raw polynomial basis expansion. Let \(X\_1 \sim \mathcal{U}(\text{range}[1], \text{range}[2])\) be the base input. A cubic transformation with vertical jitter is used to define \(X\_2 = X\_1 + X\_1^2 - X\_1^3 + \varepsilon\_2\$, where \$\varepsilon\_2 \sim \mathcal{U}(0, 0.5)\).

@@

\[
X\_j \sim \mathcal{N}(0, 0.1^2)\$ for \$j = 3, \dots, p.
\]

@@

The \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a rectangular base. The pyramid is embedded in a \(p\)-dimensional space, with a tip at height zero and base at height \(h\). The shape tapers linearly from the base dimensions \((l\_x, l\_y)\) to a smaller rectangular section with side lengths \((2r\_t, 2r\_t)\) at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

Let \(x\_1, x\_2, \dots, x\_p\) denote the coordinates of the generated points. The final dimension \(x\_p\) encodes the height of each point and is drawn from an exponential distribution capped at \(h\):

@@

Let \(r\_x(z)\) and \(r\_y(z)\) denote the half-widths of the rectangular cross-section at height \(z\):

@@

\[
x_1 \sim \mathcal{U}(-r_x(z),\ r_x(z)), \quad
x_2 \sim \mathcal{U}(-r_y(z),\ r_y(z)), \quad
x_3 \sim \mathcal{U}(-r_x(z),\ r_x(z))
\].

For \(p > 3\), the remaining \(p - 4\) dimensions (i.e., \(x\_4\) to \(x\_{p-1}\)) are tapered toward zero with decreasing height and generated as:

\[
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right), \quad \text{for } j = 4, \dots, p-1.
\]

@@

Let \(x\_1, x\_2, \dots, x\_p\) denote the coordinates of the generated points. The final coordinate \$x\_p\$ encodes the height \(z\) of each point and is drawn from an exponential distribution capped at \(h\):

@@

\[
u, v \sim \mathcal{U}(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v.
\]

@@

\[
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right).
\]

@@

Let \(x\_1, x\_2, \dots, x\_p\) denote the coordinates of the generated points. The final coordinate \(x\_p\) encodes the height \(z\) of each point and is drawn from an exponential distribution capped at \(h\):

@@

Let \(x\_1, x\_2, \dots, x\_p\) denote the coordinates of the generated points. The final coordinate \(x\_p\) encodes the height \(z\):

\[
x_p = z \sim \mathcal{U}(0, h).
\]

The radius at height \(z\) scales linearly from zero (tip) to the base radius \(r\_b\):

@@

Each point is placed within a regular hexagon in the plane \((x\_1, x\_2)\), using a randomly chosen hexagon sector angle \(\theta \in \{0, \frac{\pi}{3}, \frac{2\pi}{3}, \pi, \frac{4\pi}{3}, \frac{5\pi}{3}\}\) and a uniformly random radial scaling factor:

\[
\theta \sim \text{Uniform sample from 6 hexagon angles}, \quad
r_{\text{point}} \sim \sqrt{\mathcal{U}(0, 1)}.
\]

@@

\[
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h} \right).
\]

@@

Let \(x\_1, x\_2, \dots, x\_p\) denote the coordinates of the generated points. The generation process begins with an initial point \(T\_0 \in [0, 1]^p\) drawn from a uniform distribution:

\[
T_0 \sim \mathcal{U}(0, 1)^p.
\]

Let \(C\_1, C\_2, \dots, C\_{p+1}\) denote the corner vertices of a \(p\)-dimensional simplex. At each iteration \(i = 1, \dots, n\), a new point is computed by taking the midpoint between the previous point \(T\_{i-1}\) and a randomly selected vertex \(C\_k\):

@@

\[
X_j \sim \mathcal{N}(0, 0.05^2),\text{ for }j = 4, \dots, p.
\]

@@

In generation of \(k\) smaller uniform spheres, each sphere contains \(n_2\) points that are sampled uniformly on a sphere with a radius of \(r_2\). These spheres are positioned at distinct random locations in \(p\)-space, with the center of each sphere being drawn from a normal distribution \(\mathcal{N}(0, \texttt{loc}^2 I_p)\).

@@

\[
X_j \sim \mathcal{N}(0, 0.05^2),\text{ for }j = 5, \dots, p.
\]

@@

\[
X_j = s_j \cdot \theta + \varepsilon_j,\text{ where }\varepsilon_j \sim \mathcal{N}(0, 0.5^2),\text{ and }s_j\text{ is a random scale}.
\]

@@

Let \(\theta \sim \mathcal{U}(0, 3\pi)\) represent a random angle on a circular base and \(z \sim \mathcal{U}(0, h)\) represent the height along the cylinder.

@@

\[
X_j \sim \mathcal{N}(0, 0.05^2),\text{ for }j = 5, \ldots, p.
\]

@@

\begin{itemize}
\tightlist
\item
  \(X_1 = \sin(\phi) \cdot \cos(\theta)\)
\item
  \(X_2 = \sin(\phi) \cdot \sin(\theta)\)
\item
  \(X_3 = \cos(\phi) + \varepsilon\), where \(\varepsilon \sim \mathcal{U}(-0.5, 0.5)\) introduces vertical jitter.
\item
  \(X_4 = \theta / \max(\theta)\): a normalized progression along the spiral path.
\end{itemize}

@@

where:
\(a_j \in \{2, 3, 4, 5\}\) is a randomly chosen polynomial power,
\(s_j \sim \mathcal{U}(0.5, 2)\) is a scale factor,
\(\eta_j \sim \mathcal{U}(-\sigma_j, 2\sigma_j)\), with \(\sigma_j \sim \mathcal{U}(0, 0.05)\), adds mild randomness.

@@

\begin{itemize}
\item
  \(X_2 = \sin(\theta)\): circular trajectory along the y-axis.
\item
  \(X_3 = 0.05 \cdot \theta + \varepsilon_3\), with \(\varepsilon_3 \sim \mathcal{U}(-0.5, 0.5)\): linear progression (height) with vertical jitter, simulating a helix.
\item
  \(X_4 = 0.1 \cdot \sin(\theta)\): oscillates with \(\theta\), representing a periodic ``wobble'' along the fourth dimension.
\end{itemize}

@@

\[
X_j \sim \mathcal{N}(0, 0.05^2), \quad j = 5, \dots, p.
\]

@@

The growth pattern resembles a cone, with the height increasing according to \(X_3 = 2 \cdot \theta / \max(\theta) + \varepsilon_3\), with \(\varepsilon_3 \sim \mathcal{U}(-0.1, 0.6).\)

Spiral modulation in the fourth dimension is represented by \(X_4 = \theta \cdot \sin(2\theta) + \varepsilon_4\), with \(\varepsilon_4 \sim \mathcal{U}(-0.1, 0.6)\) which simulates a twisting helical component in a non-radial dimension.

@@

\[
X\_j \sim \mathcal{N}(0, 0.05^2), \quad j = 5, \dots, p.
\]

@@

\begin{itemize}
\tightlist
\item
  \(X_1 \sim \mathcal{U}(0.1, 2)\): base variable (avoids zero to prevent division errors).
\item
  \(X_3 \sim \mathcal{U}(0.1, 0.8)\): independent auxiliary variable.
\item
  \(X_2 = \frac{hc}{X_1} + \text{non\_fac} \cdot \sin(X_1)\): non-linear combination of \textbf{hyperbolic and sinusoidal transformations}, creating sharp curvature and oscillation.
\item
  \(X_4 = \cos(\pi \cdot X_1) + \varepsilon\), with \(\varepsilon \sim \mathcal{U}(-0.1, 0.1)\): additional nonlinear variation based on cosine, simulating more subtle periodic structure.
\end{itemize}

@@

\section{Discussion}\label{discussion}

@@

This article is created using \CRANpkg{knitr} \citep{yihui2015} and \CRANpkg{rmarkdown} \citep{yihui2018} in R with the \texttt{rjtools::rjournal\_article} template.

>>>>

The development version can be installed from GitHub:

@@

\subsection{Usage}\label{usage}

@@

The \texttt{gen\_expbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) exponential branches in \(2\text{-}D\), with optional noise dimensions to embed the structure in a \(p\)-dimensional space. These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

@@

\(X_1 \sim U(-2, 2)\)
\(X_2 = \exp(\pm s_i \cdot X_1) + \epsilon\),

where:
\(\epsilon \sim U(0, 0.1)\) adds local jitter,

\(s_i \in [0.5, 2]\) is randomly sampled for each branch.

@@

\[
X_j \sim N(0, 0.1^2), \quad j = 3, \dots, p.
\]

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from \(X_1 \sim U(-2, 8)\), \(X_2 = 0.5 \cdot X_1 + \epsilon\). The Branch 2 is generated from \(X_1 \sim U(-6, 2)\), \(X_2 = -0.5 \cdot X_1 + \epsilon\). \(\epsilon \sim U(0, 0.5)\).

Branches 3 to \(k\) are added iteratively. Each starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches.

\(X_1\) values span a short segment (\(x_{start}\) to \(x_{start} + 1\)),

\(X_2 = s_i \cdot (X_1 - x_{start}) + y_{start} + \epsilon\) where \(s_i\) is a sampled slope from a filtered set and \(\epsilon \sim U(0, 0.2)\).

@@

\[
X_j \sim N(0, 0.05^2)\text{ for }j = 3, \dots, p.
\]

@@

Branch 1 is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1 \cdot X_1 + 1 \cdot X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane.

Branch 2 is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1 \cdot X_{1} - 2 \cdot X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane.

Branches \(3\) to \(k\) are added iteratively. Each new branch begins at a randomly chosen point within a restricted horizontal band: \(X_{1} \in [-0.15, 0.15]\) (to ensure connectivity with earlier branches), spans a unit-length interval on \(X_{1}\):\(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), and has the structure:

@@

Here, \(s_i\) is a sampled scale factor (slope-like term) controlling curvature, and \((x_{\text{start}}, y_{\text{start}})\) is the starting point from the existing structure.

@@

\[
X_j \sim N(0, 0.05^2), \quad \text{for } j = 3, \dots, p.
\]

@@

To construct each branch, a unique or repeated pair of dimensions is selected from the \(\binom{p}{2}\) possible \(2\text{-}D\) combinations. If \(k \leq \binom{p}{2}\), combinations are sampled without replacement. If \(k > \binom{p}{2}\), additional pairs are sampled with replacement to reach \(k\) total branches. Each selected pair \((x_{i1}, x_{i2})\) defines the \(2\text{-}D\) plane for branch \(i\).

For branch \(i\), \(n_i\) points are generated, where \(\sum_{i=1}^k n_i = n\). The structure follows:

\[
x_{i1} \sim U(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5),
\]

where \(s_i\) is a scale factor controlling the slope. When \(k \leq \binom{p}{2}\), \(s_i = 1\). When sampling with replacement, \(s_i\) is drawn from the set \({1, 1.5, 2, \dots, 8}\) in increments of \(0.5\).

@@

\[
x_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
\]

@@

Let \(\binom{p}{2}\) denote the number of unique \(2\text{-}D\) subspace combinations. When \(k \leq \binom{p}{2}\), \(k\) distinct subspace pairs \((x_{i1}, x_{i2})\) are sampled without replacement. Otherwise, combinations are selected with replacement to reach \(k\) total branches. For each branch \(i = 1, \dots, k\), a random scale factor \(s_i\) is used to vary the curvature:

\begin{itemize}
\tightlist
\item
  If \(k \leq \binom{p}{2}\): \(s_i = 1\)
\item
  If \(k > \binom{p}{2}\): \(s_i \sim \text{Uniform sample from } {1, 1.5, \dots, 8}\)
\end{itemize}

Each branch contains \(n_i\) points such that \(\sum_{i=1}^k n_i = n\), where the vector \((n_1, \dots, n_k)\) is randomly drawn using the helper function \texttt{gen\_nsum()} to partition \(n\).

Within its assigned subspace \((x_{i1}, x_{i2})\), branch \(i\) is defined by:

\[
x_{i1} \sim U(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)
\]

This forms a smooth downward-opening parabola in the plane defined by \((x_{i1}, x_{i2})\), with the degree of curvature controlled by \(s_i\).

@@

Each variable \(X_j\) represents integer grid coordinates (e.g., 1, 2, 3, \ldots) in dimension \(j\):

\[
X_j \in {1, \dots, k_j},\text{ where }\prod_{j=1}^p k_j \approx n.
\]

@@

Points are sampled independently from \(X_1, X_2, X_3 \sim U(-0.5, 0.5)\) of a cube of side length 1. This results in a uniform cloud of points filling the unit cube in \(3\text{-}D\) space.

@@

\[
X_j \sim N(0, 0.05^2), \quad j = 4, \dots, p.
\]

@@

The initial points are drawn from a uniform cube centered at the origin: \(X_1, X_2, X_3 \sim U(-0.5, 0.5)\).

@@

Remove if \(\sqrt{X_1^2 + X_2^2 + X_3^2} \leq 0.5\).

@@

\[
X_j \sim N(0, 0.05^2), \quad j = 4, \dots, p.
\].

@@

\[
X_i \sim N_p(\boldsymbol{0}, \Sigma)
\]

@@

\(\text{scale}_i \sim U(-10, 10)\): randomly chosen direction and stretch.

\(\text{shift}_i \sim U(-300, 300)\): large offset to spread features apart.

\(\epsilon \sim N(0, 0.03 \cdot n)\): Gaussian noise scaled to dataset size to introduce mild irregularity.

@@

\[
X_1 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \cos(\theta),
\]
\[
X_2 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \sin(\theta),
\]
\[
X_3 = \frac{w}{2} \sin(\theta/2).
\]

@@

\[
x_j \sim N(0, 0.05^2), \quad j = 4, \dots, p.
\]

@@

The curve is constructed by drawing uniformly spaced inputs and applying a second-degree polynomial transformation. Let \(X_1 \sim U(\text{range}[1], \text{range}[2])\) be the independent variable. A raw polynomial basis of degree 2 is used to form \(X_2 = X_1 - X_1^2 + \varepsilon_2\), where \(\varepsilon_2 \sim U(0, 0.5)\).

@@

The shape is generated using a third-degree raw polynomial basis expansion. Let \(X_1 \sim U(\text{range}[1], \text{range}[2])\) be the base input. A cubic transformation with vertical jitter is used to define \(X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2\), where \(\varepsilon_2 \sim U(0, 0.5)\).

@@

\[
X_j \sim N(0, 0.1^2)\$ for \$j = 3, \dots, p.
\]

@@

The \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a rectangular base. The pyramid is embedded in a \(p\)-dimensional space, with a tip at height zero and base at height \(h\). The shape tapers linearly from the base dimensions \((l_x, l_y)\) to a smaller rectangular section with side lengths \((2r_t, 2r_t)\) at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

Let \(x_1, x_2, \dots, x_p\) denote the coordinates of the generated points. The final dimension \(x_p\) encodes the height of each point and is drawn from an exponential distribution capped at \(h\):

@@

Let \(r_x(z)\) and \(r_y(z)\) denote the half-widths of the rectangular cross-section at height \(z\):

@@

\[
x_1 \sim U(-r_x(z),\ r_x(z)), \quad
x_2 \sim U(-r_y(z),\ r_y(z)), \quad
x_3 \sim U(-r_x(z),\ r_x(z))
\].

For \(p > 3\), the remaining \(p - 4\) dimensions (i.e., \(x_4\) to \(x_{p-1}\)) are tapered toward zero with decreasing height and generated as:

\[
x_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right), \quad \text{for } j = 4, \dots, p-1.
\]

@@

Let \(x_1, x_2, \dots, x_p\) denote the coordinates of the generated points. The final coordinate \$x\_p\$ encodes the height \(z\) of each point and is drawn from an exponential distribution capped at \(h\):

@@

\[
u, v \sim U(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v.
\]

@@

\[
x_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right).
\]

@@

Let \(x_1, x_2, \dots, x_p\) denote the coordinates of the generated points. The final coordinate \(x_p\) encodes the height \(z\) of each point and is drawn from an exponential distribution capped at \(h\):

@@

Let \(x_1, x_2, \dots, x_p\) denote the coordinates of the generated points. The final coordinate \(x_p\) encodes the height \(z\):

\[
x_p = z \sim U(0, h).
\]

The radius at height \(z\) scales linearly from zero (tip) to the base radius \(r_b\):

@@

Each point is placed within a regular hexagon in the plane \((x_1, x_2)\), using a randomly chosen hexagon sector angle \(\theta \in \{0, \frac{\pi}{3}, \frac{2\pi}{3}, \pi, \frac{4\pi}{3}, \frac{5\pi}{3}\}\) and a uniformly random radial scaling factor:

\[
\theta \sim \text{Uniform sample from 6 hexagon angles}, \quad
r_{\text{point}} \sim \sqrt{U(0, 1)}.
\]

@@

\[
x_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h} \right).
\]

@@

Let \(x_1, x_2, \dots, x_p\) denote the coordinates of the generated points. The generation process begins with an initial point \(T_0 \in [0, 1]^p\) drawn from a uniform distribution:

\[
T_0 \sim U(0, 1)^p.
\]

Let \(C_1, C_2, \dots, C_{p+1}\) denote the corner vertices of a \(p\)-dimensional simplex. At each iteration \(i = 1, \dots, n\), a new point is computed by taking the midpoint between the previous point \(T_{i-1}\) and a randomly selected vertex \(C_k\):

@@

\[
X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p.
\]

@@

In generation of \(k\) smaller uniform spheres, each sphere contains \(n_2\) points that are sampled uniformly on a sphere with a radius of \(r_2\). These spheres are positioned at distinct random locations in \(p\)-space, with the center of each sphere being drawn from a normal distribution \(N(0, \texttt{loc}^2 I_p)\).

@@

\[
X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p.
\]

@@

\[
X_j = s_j \cdot \theta + \varepsilon_j,\text{ where }\varepsilon_j \sim N(0, 0.5^2),\text{ and }s_j\text{ is a random scale}.
\]

@@

Let \(\theta \sim U(0, 3\pi)\) represent a random angle on a circular base and \(z \sim U(0, h)\) represent the height along the cylinder.

@@

\[
X_j \sim N(0, 0.05^2),\text{ for }j = 5, \ldots, p.
\]

@@

\begin{itemize}
\tightlist
\item
  \(X_1 = \sin(\phi) \cdot \cos(\theta)\)
\item
  \(X_2 = \sin(\phi) \cdot \sin(\theta)\)
\item
  \(X_3 = \cos(\phi) + \varepsilon\), where \(\varepsilon \sim U(-0.5, 0.5)\) introduces vertical jitter.
\item
  \(X_4 = \theta / \max(\theta)\): a normalized progression along the spiral path.
\end{itemize}

@@

where:
\(a_j \in \{2, 3, 4, 5\}\) is a randomly chosen polynomial power,
\(s_j \sim U(0.5, 2)\) is a scale factor,
\(\eta_j \sim U(-\sigma_j, 2\sigma_j)\), with \(\sigma_j \sim U(0, 0.05)\), adds mild randomness.

@@

\begin{itemize}
\item
  \(X_2 = \sin(\theta)\): circular trajectory along the y-axis.
\item
  \(X_3 = 0.05 \cdot \theta + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.5, 0.5)\): linear progression (height) with vertical jitter, simulating a helix.
\item
  \(X_4 = 0.1 \cdot \sin(\theta)\): oscillates with \(\theta\), representing a periodic ``wobble'' along the fourth dimension.
\end{itemize}

@@

\[
X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p.
\]

@@

The growth pattern resembles a cone, with the height increasing according to \(X_3 = 2 \cdot \theta / \max(\theta) + \varepsilon_3\), with \(\varepsilon_3 \sim U(-0.1, 0.6).\)

Spiral modulation in the fourth dimension is represented by \(X_4 = \theta \cdot \sin(2\theta) + \varepsilon_4\), with \(\varepsilon_4 \sim U(-0.1, 0.6)\) which simulates a twisting helical component in a non-radial dimension.

@@

\[
X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p.
\]

@@

\(X_{1} \sim U(0.1, 2)\): base variable (avoids zero to prevent division errors).
\(X_{3} \sim U(0.1, 0.8)\): independent auxiliary variable.

:non-linear combination of \textbf{hyperbolic and sinusoidal transformations}, creating sharp curvature and oscillation.
\(X_{4} = \cos(\pi \cdot X_{1}) + \varepsilon\), with \(\varepsilon \sim U(-0.1, 0.1)\): additional nonlinear variation based on cosine, simulating more subtle periodic structure.

@@

\section{Conclusion}\label{conclusion}

@@

These \texttt{R} packages were used for this work: \texttt{cli} \citep{gabor2025}, \texttt{tibble} \citep{kirill2023}, \texttt{gtools} \citep{gregory2023}, \texttt{dplyr} \citep{hadley2023}, \texttt{stats} \citep{core2025}, \texttt{tidyr} \citep{hadley2024}, \texttt{purrr} \citep{hadley2025}, \texttt{mvtnorm} \citep{alan2009}, \texttt{geozoo} \citep{barret2016}, and \texttt{MASS} \citep{venables2002}. This article is created using \CRANpkg{knitr} \citep{yihui2015} and \CRANpkg{rmarkdown} \citep{yihui2018} in R with the \texttt{rjtools::rjournal\_article} template.
